The code generator presented here accepts as input a sequence of stencil
statements that are expressed in a restricted subset of C consistent with the
grammar described in figure \ref{fig:start-grammar}. The input to the code
generator is a full stencil code in a .cu file, with the regions that the user
wants to reorder annotated with \texttt{\#pragma begin stencil$\ldots$} and a
matching \texttt{\#pragma end}. The parser is simplified and does not perform
any syntactic verification, so when creating a new benchmark, it is the user's
responsibility to ensure that the syntax adheres to the specifications for the
code generator to work correctly. 

In the grammar of figure \ref{fig:start-grammar}, a $\mathit{BaseExpr}$ in a
statement represents an accessed storage location, and is identified by a
\textit{label}. A crucial part of the reordering framework is to identify the
\textit{same} $\mathit{BaseExpr}$. This is done by checking the equivalence of
the labels that are assigned to any two $\mathit{BaseExpr}$. This is
complicated by the fact that post unrolling, one has to guarantee that
\texttt{A[i+1+1]} will be assigned the same label as \texttt{A[i+2]}. In
general, the array index expressions can be rewritten in multiple, albeit
equivalent ways -- \texttt{A[i+j]} is the same as \texttt{A[j+i]},
\texttt{A[i+1+j]} is the same as \texttt{A[i+j+1]}, and so on. To ensure that
the semantically equivalent $\mathit{BaseExpr}$ are assigned the same label,
the code generator has a mechanism that ``canonicalizes'' the accesses, so that
semantically equivalent accesses have the same pattern. We define an array index expression
to comprise of an iterator-access component (e.g., \texttt{(i+j)} in access
\texttt{A[i+j+2]}), and an integer offset component (e.g., 2 in accesses
\texttt{A[i+j+1+1]} and \texttt{A[i+j+2]}). Within the code generator, this
logic is encoded in the function \textit{array\_access\_info ()}, and it works
in two steps: 
\begin{compactenum}
\item it separates out iterator-access component and the integer offset from an index expression
\item it canonicalizes both the extracted components. For the iterator-access
component, this involves identifying the presence of an operator in the
iterator-access, and then exploiting the properties of the operator to align
the iterators in lexicographical order. For example, \texttt{(j-i)} will be
converted to \texttt{(j+(-i))}, and then \texttt{(j \textit{op} i)} will be
rearranged as \texttt{(i \textit{op} j)}, if \texttt{i} precedes \texttt{j} in
lexicographical ordering, and \textit{op} is commutative, like +,*. For the
offset component, it identifies the operators involved, and performs the
mathematical operation to reduce the offset component to a single integer
value.
\end{compactenum}
The canonicalization part is non-trivial, and currently primitive in
implementation in the code generator. We currently restrict the number of
operators in the iterator-access component to 1. For example, we allow
A[k+1+1][2*j+2][i] since we can canonicalize each of the array index
expressions, but not something as complex as A[(k+2)*(k+3)][j][i], since this
access will result in an iterator-offset component (k*k+5k) where there are two
arithmetic operations involved.
 

In summary, We impose the following restrictions on the statements:
\begin{itemize}
\item The statements must not modify the iterators (e.g., i, j and k in listing
\ref{listing:new-eg}) in any way.  
\item The statements can only have mathematical functions like sine, sqrt,
cosine, etc. on the RHS, that are side-effect free.  
\item The accessed locations in the stencil statements are either scalars or
array elements. The array index expressions can only comprise the loop
iterators, integers or simple arithmetic operations between them (+,-,*,/).
We allow only \textit{one} arithmetic operation in the iterator-access component
of the array index expressions, and the offset component must be an integer.
\end{itemize} 

Once the parsing is complete, the code
generator lowers the input further to sub-statements consistent with the
grammar of figure \ref{fig:lower-grammar}. Informally, this lowering is like
GIMPLE IR of GCC \cite{gcc} with accumulations. 


\begin{figure}[t!]    
\footnotesize
  \begin{align}
    \hline
    \langle\text{Program}\rangle ::=& \;  \langle\text{StmtList}\rangle {\color{tablecolor!60}\texttt{;}} \nonumber \\
    \langle\text{StmtList}\rangle ::=& \; \langle\text{Stmt}\rangle \langle\text{StmtList}\rangle \; | \; \langle\text{Stmt}\rangle \nonumber \\
    \langle\text{Stmt}\rangle ::=& \; \langle\text{BaseExpr}\rangle \langle\text{AssignOp}\rangle \langle\text{Expr}\rangle {\color{tablecolor!60}\texttt{;}} \nonumber \\
    \langle\text{Expr}\rangle ::=& \; \text{Side-effect-free expression of~} \langle\text{BaseExpr}\rangle \nonumber \\
    \langle\text{BaseExpr}\rangle ::=& \; \langle\text{ID}\rangle \; | \; \langle\text{ArrayExpr}\rangle \nonumber \\
    \langle\text{ArrayExpr}\rangle ::=& \; \langle\text{ID}\rangle \langle\text{ArrayDims}\rangle \nonumber \\
    \langle\text{ArrayDims}\rangle ::=& \; {\color{tablecolor!60}\texttt{[}}\langle\text{AccessExpr}\rangle {\color{tablecolor!60}\texttt{]}} \langle\text{ArrayDims}\rangle \; | \; {\color{tablecolor!60}\texttt{[}}\langle\text{AccessExpr}\rangle {\color{tablecolor!60}\texttt{]}} \nonumber \\
    \langle\text{AccessExpr}\rangle ::=& \; \text{Side-effect-free expression of~} \langle\text{AccessElement}\rangle \text{ with restrictions described in text}\nonumber \\
	\langle\text{AccessElement}\rangle ::=& \; \langle\text{Iterator}\rangle \; | \; \langle\text{Const}\rangle \nonumber \\
    \langle\text{AssignOp}\rangle ::=& \; {\color{tablecolor!60}\texttt{=}} \; | \; {\color{tablecolor!60}\texttt{+=}} \; | \; {\color{tablecolor!60}\texttt{-=}} \; | \; {\color{tablecolor!60}\texttt{*=}} \; | \; \ldots \nonumber  \\
    \langle\text{DataType}\rangle ::=& \; {\color{tablecolor!60}\texttt{double}} \; | \; {\color{tablecolor!60}\texttt{float}} \; | \; {\color{tablecolor!60}\texttt{int}} \; | \; {\color{tablecolor!60}\texttt{bool}} \; | \; \ldots \nonumber  \\
    \hline \nonumber
  \end{align}
\vspace*{-0.6cm}
  \caption{Grammar for valid input statements to the code generator}
\label{fig:start-grammar}
\end{figure}


\begin{figure}[t!]
  \footnotesize
  \begin{align}
    \hline
    \langle\text{CodeBody}\rangle ::=& \;  \langle\text{StmtList$_l$}\rangle {\color{tablecolor!60}\texttt{;}} \nonumber \\
    \langle\text{StmtList$_l$}\rangle ::=& \; \langle\text{Assignment$_l$}\rangle {\color{tablecolor!60}\texttt{;}} \langle\text{StmtList$_l$}\rangle \; | \; \langle\text{Assignment$_l$}\rangle \nonumber \\
    \langle\text{Assignment$_l$}\rangle ::=& \; \langle\text{BaseExpr}\rangle \langle\text{AssignOp}\rangle \langle\text{Expr$_l$}\rangle \nonumber \\
    \langle\text{Expr$_l$}\rangle ::=& \; \langle\text{BaseExpr}\rangle \langle\text{BinaryOp$_l$}\rangle \langle\text{BaseExpr}\rangle \; | \; \langle\text{BaseExpr}\rangle \nonumber \\
    \langle\text{BinaryOp$_l$}\rangle ::=& \; {\color{tablecolor!60}\texttt{+}} \; | \; {\color{tablecolor!60}\texttt{-}} \; | \; {\color{tablecolor!60}\texttt{*}} \; | \; {\color{tablecolor!60}\texttt{/}} \; | \; {\color{tablecolor!60}\texttt{|}} \; | \; {\color{tablecolor!60}\texttt{\&}} \; | \; \ldots \nonumber  \\
    \hline \nonumber
  \end{align}
   \vspace*{-0.6cm}
  \caption{Grammar for the lowered 3-address IR for instruction reordering}
\label{fig:lower-grammar}
\end{figure}



\section{Creating an example}
Listing \ref{listing:new-eg} shows a snippet of a cuda input code that can be
passed as an input to the reordering framework. The computation in lines 5-7 is
affine, and in compliance with the restrictions we placed on the input grammar.
Note the pragmas in line 4 and 8 that surround the statement -- these demarcate
the region that the code generator can parse and optimize. Additionally, the
pragma supplies the unrolling factor along each dimension. Given an input file
with several such pragma-demarcated regions, the steps in optimizing the code
will be: 

\begin{enumerate} 
\item A script will remove the statements between all the pragma, and put them
into separate \texttt{.idsl} files, each named based on the stencil name (e.g.,
\texttt{stencil1.idsl} for the example in listing \ref{listing:new-eg}).
Therefore, each stencil name must be distinct.
\item From the pool of \texttt{stencilname.idsl} files generated, the code
generator will pick one at a time, parse it, optimize it, and write the results
back in \texttt{stencilname.cu}.  
\item A script will then read the contents from \texttt{stencilname.cu}, and
replace the original statements demarcated by \texttt{\#pragma begin
stencilname $\ldots$ \#pragma end stencilname} with the optimized code. 
\end{enumerate} 
The changes are made in-place to the input cuda file.
Please look at the benchmarks in the examples folder to learn more about how to use this
code generator for your applications. 
 
\begin{figure}[t!]
\vspace*{-1em}
%\vspace*{-15pt}
\begin{lstlisting}[escapechar=@,caption={The input representation with pragmas},label={listing:new-eg}]
  for (k=1; k<=N-2; k++) {
       for (j=1; j<=N-2; j++) {
           for (i=1; i<=N-2; i++) {
#pragma begin stencil1 unroll k=4,j=1,i=1
                   out[k][j][i] = $a$*(in[k+1][j][i]) + $b$*(in[k][j-1][i] +
                              in[k][j][i-1] + in[k][j][i] + in[k][j][i+1] +
                                  in[k][j+1][i]) + $c$*(in[k-1][j][i]);
#pragma end stencil1
           }
       }
  }
\end{lstlisting}
\end{figure}



